### 2017-01-28 thesis/methods/methods.tex
Finished Creighton obesity metagene section draft

In order to ertain set of genes are up or down regulated by a sample, some sort of score based on these genes must be calculated.
\Gls{svd} is a mathematical method that splits a matrix into several matrices that represent the original matrix \citep{Golub1970}.
With \gls{svd}, a matrix $X$ of size $n$ genes by $m$ arrays (or samples) can be represented as:
\begin{equation}
	\label{eq:svd}
	X = UDV^T
\end{equation}
where the columns of $U$ and $V$ contain the left and right singular vectors of $X$ respectively, and $D$ contains the singular values of $X$.

@@

With respect to this project, \gls{svd} was used to assess whether the metagene created from various obesity or pathway associated genes were significantly associated with certain clinical variables, such as \gls{bmi}.
To do this, the normalised original expression data was reduced in size so that it only included the genes of ineterest (either obesity or pathway associated) and the samples, then \texttt{svd} function from the \textit{base} package in \gls{r} was used to apply \gls{svd} to the matrix.
The first principal component was taken from the $V$ matrix and was used as the metagene scores for the samples in the data.

>>>>

In order to determine whether a set of genes are up or down regulated by a sample, some sort of score based on these genes must be calculated.
\Gls{svd} is a mathematical method that splits a matrix into several matrices that represent the original matrix \citep{Golub1970}.
With \gls{svd}, a matrix $X$ of size $n$ genes by $m$ arrays (or samples) can be represented as:
\begin{equation}
	\label{eq:svd}
	X = UDV^T
\end{equation}
where the columns of $U$ and $V$ contain the left and right singular vectors of $X$ respectively, and $D$ contains the singular values of $X$.

@@

With respect to this project, \gls{svd} was used to assess whether the metagene created from various obesity or pathway associated genes were significantly associated with certain clinical variables, such as \gls{bmi}.
To do this, the normalised original expression data was reduced in size so that it only included the genes of interest (either obesity or pathway associated) and the samples, then \texttt{svd} function from the \textit{base} package in \gls{r} was used to apply \gls{svd} to the matrix.
The first principal component was taken from the $V$ matrix and was used as the metagene scores for the samples in the data.

### 2017-01-28 thesis/results/results.tex
Finished Creighton obesity metagene section draft

In this chapter, the relationship between the obesity associated genetic signatures from previous studies and different cancer types were investigated; new obesity associated genetic signatures were identified in the data set from \citet{Creighton2012} study; and the presence of common genes or pathways across different cancer types was examined.

@@

% \begin{figure}[h!]
% 	\centering
% 	\includegraphics[width=0.8\linewidth]{intro/hallmarks}
% 	\caption[Hallmarks of Cancer]{Hallmarks of Cancer. The ten Hallmarks of Cancer as proposed by Hanahan and Weinberg in 2011. (Figure adapted from \citet{Hanahan2011})}
% 	\label{fig:hallmarks}
% \end{figure}

@@

@@

@@

>>>>

The obesity associated genetic signatures are central to this project in order to clarify the relationship between \gls{bmi} and cancer.
Firstly in this chapter, the previously identified obesity associated signatures from the studies conducted by \citet{Creighton2012} and \citet{Fuentes-Mattei2014} were examined in turn to judge the agreement of these signatures with the sample \gls{bmi}, as presented in thier results.
Secondly, novel obesity associated genetic signatures were identified in the \citet{Creighton2012} data set and was compared with the signatures.
Lastly, the presence of common genes or pathways that were associated with obesity in multiple types of cancer was explored using the \gls{icgc} cancer data sets.

@@

Obesity metagene was created using the obesity associated genetic signatures from \citet{Creighton2012} study.
Since there was no description about the normalisation method used by Creighton's group, the \gls{rma} method was used to  normalise the Creighton data set.
The obesity metagene was plotted in a heatmap with the sample gene expression (\cref{fig:crmetaheat1}) to check whether the metagene scores were in accordance with the overall gene expression of the samples.

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.8\linewidth]{results1/crmeta_original}
	\caption[Heatmap of Creighton's metagene in Creighton's data]{Heatmap showing Creighton's obesity associated metagene with the gene expressions of obesity associated genes in Creighton's data. Level of expression is represented in the top right histogram, where low and high gene expression were colour-coded with blue and red, respectively. Each row of the heatmap represents a gene from the obesity associated genetic signature, and each column of the heatmap represents a sample from Creighton's data. The obesity associated metagene scores of the samples are shown in a separate row at top of the heatmap, and the tree diagram of the heirarchical clustering of the genes is shown to the left of the heatmap. For clarity, the metagene scores were flipped in order to match the results from \citet{Creighton2012} paper.}
	\label{fig:crmetaheat1}
\end{figure}

As shown in \cref{fig:crmetaheat1}, high obesity associated metagene score of the sample reflected low expression in majority of the genes in the signature, and in contrast, low obesity associated metagene score of the sample reflected high expression in majority of the genes in the signature.
This was consistent with the reported property of the obesity associated signature by \citet{Creighton2012}.
To provide further evidence that the obesity associated metagenes were in fact associated with the \gls{bmi} status and \gls{bmi} value of the samples, a box plot and a scatter plot were created, respectively (\cref{fig:crmetaboxplot1}).

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\hfill
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\caption[Creighton metagene and sample \gls{bmi}/\gls{bmi} status]{(insert figure of box plot and scatter plot)}
	\label{fig:crmetaboxplot1}
\end{figure}

\cref{fig:crmetaboxplot1} clearly showed that the obesity associated metagene from \citet{Creighton2012} study significantly associated with the sample \gls{bmi} status, as well as the \gls{bmi} value of the samples (include p-value and r-squared).
It should be noted here that the obesity associated metagene significantly associated with the samples that were obese, but not the samples that were overweight.
This was due to the fact that the obesity associated genetic signatures were originally identified from the comparison of the samples that were obese with the samples that were not obese, and therefore the metagene scores were significant with obese group, but not the overweight group.
\\

\noindent
Now that the association of the obesity metagene from \citet{Creighton2012} study was established in Creighton's data set, the obesity associated signature was transferred to the \gls{icgc} cancer data.
The direction of the obesity associated metagene was checked in the Creighton's data, so that high metagene scores reflected high sample \gls{bmi} and low metagene scores reflected low sample \gls{bmi} (\cref{fig:crmetaboxplot1}).
The transformation matrix was created in Creighton's data, as described in \cref{sub:svd}.

All of the \gls{icgc} data were normalised as described in \cref{ssub:rna_seq_data}.
Before the transformation matrix was applied to the log$_{10}$-normalised cancer data, the suitability of standardised data or untouched (non-standardised) data was determined (appendix).
From these results, the standardised data was the most suitable to use for matrix transformation.
The transformation matrix was applied to each cancer data set in turn to obtain the obesity metagenes for each of the data set.
Each obesity metagenes were plotted on a heatmap with the corresponding data set in which the metagene was taken from \cref{fig:crmetaicgc1}.
These heatmaps confirmed that the obesity associated metagene derived from Creighton's data was able to capture the overall gene expression pattern, where the metagene scores reflected the expression levels of majority of the genes in the signature.

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\hfill
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\caption[Creighton metagene and \acrshort{icgc} cancer data]{(insert ICGC heatmaps)}
	\label{fig:crmetaicgc1}
\end{figure}

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\hfill
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\caption[Creighton metagene and sample \gls{bmi}/\gls{bmi} status in \acrshort{icgc} data]{(insert figure of box plot and scatter plot for ICGC data)}
	\label{fig:crmetaboxicgc1}
\end{figure}

As before, association of the obesity metagene with the sample \gls{bmi} and \gls{bmi} status was checked in their respective cancer data set (\cref{fig:crmetaboxicgc1}).
In contrast to the strong association of the obesity metagene with sample \gls{bmi} and \gls{bmi} status in Creighton's data, the obesity metagene taken from the \gls{icgc} cancer data did not significantly associate with the sample \gls{bmi} or \gls{bmi} status in any of the \gls{icgc} cancer data.

There could be several reasons for the apparent lack of association of the metagene with the sample \gls{bmi} or \gls{bmi} status.
First, transformation matrix was derived from Creighton's microarray data, but the \gls{icgc} cancer data were \gls{rnaseq} data.
Though the log$_{10}$ normalisation and standardisation of the data was the most appropriate adjustment to be made to the \gls{rnaseq} data, this adjustment may not be equivalent to the \gls{rma} normalisation method that was used on the microarray data.
Secondly, none of the \gls{icgc} cancer was from breast as in Creighton's data.
Since the obesity associated signature was identified in breast cancer data, the signature may be specific to breast cancer and may not be suitable in other cancer types.
\\

\noindent
To check whether the metagene was specific to breast cancer microarray data, the same transformation matrix was applied to Print's breast cancer microarray data.
Print's data was normalised with \gls{rma} method as described in \cref{ssub:microarray_data}, and the transformation matrix was applied to the normalised data to obtain the metagene.
The metagene was again compared with the gene expression of the samples with a heatmap (\cref{fig:crmetaprint1}), and the association of the metagene with the sample \gls{bmi} and \gls{bmi} status was examined with box and scatter plots (\cref{fig:crmetaboxprint1}).

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\hfill
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\caption[Creighton metagene and Print's cancer data]{(insert Print heatmaps)}
	\label{fig:crmetaprint1}
\end{figure}

\begin{figure}[htp!]
	\centering
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\hfill
	\includegraphics[width=0.45\linewidth]{results1/crmeta_original}
	\caption[Creighton metagene and sample \gls{bmi}/\gls{bmi} status in Print's data]{(insert figure of box plot and scatter plot for Print data)}
	\label{fig:crmetaboxprint1}
\end{figure}

The obesity metagene managed to reflect the overall gene expression of the samples in Print's data.
However, like in \gls{icgc} cancer data, the obesity metagene scores did not significantly associate with the sample \gls{bmi} or \gls{bmi} status.
These results confirmed that the lack of association of the obesity metagene from Creighton's data was not due to the technology in which the data was gathered (mcroarray or \gls{rnaseq}), nor the type of cancer in which the transformation matrix was applied to.

All together, all of these results suggest that the obesity metagene identified in the study conducted by \citet{Creighton2012} associated with sample \gls{bmi} and \gls{bmi} status in Creighton's data, but did not transfer well in other cancer data sets.
The lack of association with \gls{bmi} was not due to the type of technology platform in which the data was gathered, as neither the \gls{icgc} \gls{rnaseq} data nor Print's microarray data did not show significant association with the obesity metagene.
Furthermore, Creighton's metagene was not dependent on the cancer type since the obesity associated genetic signature did not show significant association in neither the \gls{icgc} cancer data nor in Print's breast cancer data.

One possible reason why Creighton's obesity metagene did not show significant association with other data sets could be because the genetic signature was in fact not an obesity specific signature, but a signature that detected another clinical variable.
Another reason for this apparent lack of association could be that the genetic signature was too specific to the data and was not a broad obesity associated genetic signature, but an obesity associated signature specifically for Creighton's data set.

@@

It was surprising that Creighton's metagene did not show significant association with \gls{bmi} in other data sets.

@@

Both the metagenes generated from Creighton's data and Fuentes-Mattei's data were not associated with sample \gls{bmi}.

@@

It was clear that the metagenes created from obesity signatures were not consistent across different data sets.
To identify genes that were associated with sample \gls{bmi} in multiple cancer types, \glspl{deg} were searched (?).

### 2017-01-28 thesis/thesis.tex
tmp commit
Finished Obesity and cancer section draft
Minor edit to the table of contents and glossary style
Finished Creighton obesity metagene section draft

% \documentclass[a4paper, 12pt, oneside]{report} % for final draft compilation
\documentclass[a4paper, 12pt, oneside, draft]{report} % for draft compilation

% Preamble:
\usepackage[T1]{fontenc}                            % package for specifying output font encoding
\usepackage[utf8]{inputenc}                         % package for specifying input font encoding
\usepackage{natbib}                                 % package for reference management
\usepackage{graphicx}                               % package for graphics handling (i.e. figure images)
\usepackage{mathptmx}                               % package for Times New Roman font
\usepackage{chngcntr}                               % package for making continuous figure numbering, instead of section-based numbering
\usepackage{setspace}                               % package for setting line space (used in title page)
\usepackage[hidelinks, colorlinks]{hyperref}        % package for hyperlinking references - hidelinks option removes the ugly boxes around the huperlinks
\usepackage[md]{titlesec}                           % package for tweaking the chapter/section/subsection titles format
\usepackage[acronym, toc]{glossaries}               % package for making glossaries
\usepackage{xcolor,colortbl}                        % package for making colours
\usepackage[noabbrev,capitalise]{cleveref}          % package for referencing multiple referencing in a single \ref command
\usepackage[font=footnotesize,format=hang]{caption} % package for formatting figure caption

@@

@@

% Specify graphics path:
\graphicspath{ {./images/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% If you want to compile specific files, uncomment the line below and add the filenames in the {} :
%\includeonly{}

@@

\printglossaries

@@

% \listoftables

% \clearpage

@@

% Print out the bibliography:
\bibliography{MSc}

>>>>

\documentclass[a4paper, 12pt, oneside]{report} % for final draft compilation
% \documentclass[a4paper, 12pt, oneside, draft]{report} % for draft compilation

% Preamble:
\usepackage[T1]{fontenc}                               % package for specifying output font encoding
\usepackage[utf8]{inputenc}                            % package for specifying input font encoding
\usepackage{natbib}                                    % package for reference management
\usepackage{graphicx}                                  % package for graphics handling (i.e. figure images)
\usepackage{mathptmx}                                  % package for Times New Roman font
\usepackage{chngcntr}                                  % package for making continuous figure numbering, instead of section-based numbering
\usepackage[flushleft]{threeparttable}                 % package to add notes at the bottom of the table
\usepackage{setspace}                                  % package for setting line space (used in title page)
\usepackage[hidelinks, colorlinks]{hyperref}           % package for hyperlinking references - hidelinks option removes the ugly boxes around the huperlinks
\usepackage[md]{titlesec}                              % package for tweaking the chapter/section/subsection titles format
\usepackage[acronym]{glossaries} % package for making glossaries
\usepackage{xcolor,colortbl}                           % package for making colours
\usepackage[noabbrev,capitalise]{cleveref}             % package for referencing multiple referencing in a single \ref command
\usepackage[font=footnotesize,format=hang]{caption}    % package for formatting figure caption
\usepackage{subcaption}    % package for formatting figure caption

@@

% Change the title of Table of Contents:
\renewcommand{\contentsname}{Table of Contents}

% Change the title of Table of Contents:
\renewcommand{\thesubfigure}{\Alph{subfigure}}

@@

% Specify graphics path:
\graphicspath{{./images/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% If you want to compile specific files, uncomment the line below and add the filenames in the {} :
\includeonly{./results/results}

@@

\addcontentsline{toc}{section}{Glossary}

\printglossary[style=altlist]

@@

\addcontentsline{toc}{section}{Acronyms}

\printglossary[type=\acronymtype,style=super]

@@

\addcontentsline{toc}{chapter}{\bibname}

% Print out the bibliography:
\bibliography{./MSc,./MSc2}
